---
title: "R packages"
author: "Jeff Leek"
date: "December 14, 2015"
output: ioslides_presentation
---


## Writing R packages

* Here are some links from people who are pros
  * [Hadley's master class](http://courses.had.co.nz/12-devtools/)
  * [Creating R packages tutorial](http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf)
  * [Writing R extensions](http://cran.r-project.org/doc/manuals/R-exts.html)
* Coding style
  * [Bioconductor](http://www.bioconductor.org/developers/package-guidelines/)
  * [Hadley Wickham's version](https://github.com/hadley/devtools/wiki/Style)
  * [Google's R style guide](http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html)
  * [Roger's checklist](http://simplystatistics.org/2015/10/14/minimal-r-package-check-list/)
  * [Jeff's guide](https://github.com/jtleek/rpackages)



## Why write an R package? {.smaller}

* [If there is no code, there is no paper](http://simplystatistics.org/2013/01/23/statisticians-and-computer-scientists-if-there-is-no-code-there-is-no-paper/)
* Getting noticed (see e.g. [Yihui](http://yihui.name/))
* Getting cited (see e.g. [RMA](http://www.ncbi.nlm.nih.gov/pubmed/12925520), [sva](http://www.plosgenetics.org/article/info:doi/10.1371/journal.pgen.0030161) )
* To avoid the [statistics boogeyman](http://odin.mdacc.tmc.edu/~kabaggerly/)
* Because it is when you really learn how/why your method works
* To put together a set of functions that implement a method, or [are useful to you](http://hilaryparker.com/2013/04/03/personal-r-packages/), or summarize a body of work (like a dissertation)



## First step: naming

* Make it googleable 
* Be a little cute, but not too much
* Avoid caps 
* Package name must start with a letter
* r"method" or r"technology" is a good place to start, see e.g. [rOpenSci](http://ropensci.org/packages/index.html)
* Never let Rafa name your method ("Succs", "BumpHunter", "DER Finder")



## If you want your software to be used

<center> Documentation > Usability > Speed > Statistical superiority</center>

<img class=center src=https://raw.githubusercontent.com/jtleek/jhsph753and4/master/assets/img/homer.jpg height=350/>




## Preliminaries: Version control

[What is version control](http://git-scm.com/video/what-is-version-control)

* [Github](https://github.com/) - where the cool nerds are
* [Bitbucket](https://bitbucket.org/) - where the real nerds are
* [svn](http://subversion.tigris.org/) - where the old nerds are
* Files on your desktop - where the frustrated nerds are


## How you do this

* Step 1: Create repo on computer
* Step 2: Create repo with same name on Github
* Step 3: Follow [these instructions](https://help.github.com/articles/adding-an-existing-project-to-github-using-the-command-line/)

## Preliminaries: repositories

* [CRAN](http://cran.us.r-project.org/)
    * All types of packages
    * Minimal QC
* [Bioconductor](http://www.bioconductor.org/)
    * Mostly genomics packages
    * Peer review/high QC
* [Github](https://github.com/)
    * Any type 
    * No QC

## An important point


<img class=center src=http://simplystatistics.org/wp-content/uploads/2015/11/Screen-Shot-2015-11-06-at-1.25.01-PM.png height=250/>


http://simplystatistics.org/2015/11/06/how-i-decide-when-to-trust-an-r-package/


## Getting started - load packages


```{r , eval=FALSE}
install.packages("devtools")
install.packages("rogxygen2")
install.packages("textthat")
```

```{r load, warning=F, message=F}
library(devtools); library(roxygen2);library(testthat)
```


## Create a package

```{r ,eval=FALSE}
# Set working directory
setwd("~/Desktop")

# Create a new package
create("fdrreg")
```

<center> Add stuff to package now</center>



## R code

* Just like you would usually write
* Goes in the R/ directory
* Should be multiple files - one per major "method" at minimum
* Documentation included in files makes it easier



## Documenting R code with roxygen2


@param - Inputs

@examples - examples

@author - who wrote the function

@seealso - refers to other packages

@return - what it returns

[Here it is in practice](https://github.com/jtleek/sva-devel/blob/master/R/sva.R)

See more [here](http://courses.had.co.nz/12-devtools/slides/8-documentation.pdf)




## How to use this

```{r,eval=FALSE}
document("pkg")
check_doc("pkg")
dev_help("function")
```



## Testing R code with testthat {.smaller}

`devtools::use_testthat("mypkg")`

__Use expect_that to define subtests__

expect_that(1,equals(1))

__Example tests__

is_truth

is_false

equals

matches

prints_text

__Then arrange together__

test_that("test name", {
 expect_that(1,equals(1))
})

See more [here](http://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf)



## How to use it

* Create tests in inst/tests
* Each file can test one component
* Start file with _context("Name for set of tests")_
* To test run:
  * test_file()
  * test_dir()
* Or you can do it when running R CMD CHECK with a file tests/run_all.R with
```{r,eval=FALSE}
library(testthat)
library(mypackage)
test_package("mypackage")
```



## Vignettes versus README files

* Vignettes 
    * Go in vignettes
    * Are written in knitr
    * Are required for Bioconductor
* README
    * Can be written in markdown
    * Often useful if Github is where you will put your documentation
* [Markdown vignettes](http://yihui.name/en/2012/09/r-package-markdown-vignettes/)

[Here] is an example of some vignettes


## To use markdown vignettes

* `devtools::use_vignette("my-vignette")`
* Put them in the vignettes folder
* Add knitr to "Suggests" in DESCRIPTION
* Add line "VignetteBuilder: knitr" to DESCRIPTION 

## Releasing package

* On Github you can just post the package then tell users to do

```{r,eval=FALSE}
library(devtools)
install_github("username/package")
```

* Or for CRAN use 
```{r,eval=FALSE}
release()
```

* Or for BioC see [here](http://www.bioconductor.org/developers/)

More on this [here](http://courses.had.co.nz/12-devtools/slides/10-release.pdf)

## Adding data

* Processed data in rdas goes in _/data_
    * If you are in the working directory you can do `devtools::use_data(dataset)`
* Raw data goes in _inst/extdata_
    * You can load this with e.g. `system.file("extdata", "2012.csv", package = "testdat")`
* CRAN/Bioc have limits on data sizes. You need to look them up before submitting

[You can find more here](http://r-pkgs.had.co.nz/data.html)


## Documenting a data set


<img class=center src=dataset.png height=350/>


## Loading code to play with it

* You don't have to recomplie the package every time
* Instead use `load_all("mypkg")`
* You can also use `build("mypkg")` to build the package

## Adding rcpp code

`devtools::use_rcpp()`

<img class=center src=cpp.png height=350/>

http://r-pkgs.had.co.nz/src.html

## Adding citations

Add this to `inst/CITATION`


<img class=center src=citations.png height=350/>



## Documenting an analysis as an R package

* Put data in _data_ or _inst/extdata_
* Put analysis code in _vignettes_
* Put any R functions for helping in _R_
* Set dependencies in _DESCRIPTION_
* Post to Github
* [Read this](https://github.com/ropensci/unconf/issues/31) [and this](https://github.com/ropensci/rrrpkg) for more discussion 
* make might be better? 


