---
title: "Version control"
author: "Jeff Leek"
date: "September 2, 2015"
output: ioslides_presentation
---

## An exciting result

<img class=center src=./potti.png height=450>

http://www.nature.com/nm/journal/v12/n11/full/nm1491.html


## An exciting result

<img class=center src=./baggerly.png height=450>

https://projecteuclid.org/euclid.aoas/1267453942

## A timeline of events

<img class=center src=./timeline.png height=450>

http://www.nature.com/news/2011/110111/full/469139a/box/1.html

## Major fallout

<img class=center src=./lawsuit.png height=450>

http://dig.abclocal.go.com/wtvd/docs/Duke_lawsuit_090811.pdf
http://www.dukechronicle.com/articles/2015/05/03/duke-lawsuit-involving-cancer-patients-linked-anil-potti-settled

## An interesting talk

<img class=center src=./baggerly-talk.png height=450>

http://www.birs.ca/events/2013/5-day-workshops/13w5083/videos/watch/201308141121-Baggerly.mp4

## R markdown

<img class=center src=./rmarkdown.png height=450>

http://www.rstudio.com/ide/docs/authoring/using_markdown

## Paths

Do this: 
```{r,eval=FALSE}
setwd("../data")
setwd("./files")

setwd("..\tmp")
```

Don't do this:
```{r,eval=FALSE}
setwd("/Users/jtleek/data")
setwd("~/Desktop/files/data")

setwd("C:\\Users\\Andrew\\Downloads")
```

## Data via the web

```{r,eval=FALSE}
if(!file.exists("data")){
  dir.create("data")
}

fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/
rows.csv?accessType=DOWNLOAD"
download.file(fileUrl,
    destfile="./data/cameras.csv",
    method="curl")
list.files("./data")
dateDownloaded <- date()
```


## Things you need to be careful with

* Setting seeds for reproducible research `set.seed(1)`
* Version numbers `devtools::session_info()`, [packrat](https://rstudio.github.io/packrat/), 
* Caching and removing the cache 
* Big analyses [BiocParallel](http://bioconductor.org/packages/release/bioc/html/BiocParallel.html)


## Averge trustworthiness

<img class=center src=./trustworthy.png height=150>

## Just no

<img class=center src=http://www.phdcomics.com/comics/archive/phd101212s.gif height=450>

http://www.phdcomics.com/comics/archive/phd101212s.gif

## Just no

<img class=center src=https://imgs.xkcd.com/comics/documents.png height=450>

https://imgs.xkcd.com/comics/documents.png

## An unfortunate truth

<img class=center src=./madmen.png height=250>

https://twitter.com/mtrc/status/617075570761965568

## Another unfortunate truth

"Your closest collaborator is you six months ago, but you don't reply to emails."

http://kbroman.org/Tools4RR/assets/lectures/06_org_eda_withnotes.pdf


## And one more unfortunate truth

"Could you just re-run all that code with the [latest/different/best] parameters?"

--Every collaborator ever

http://kbroman.org/Tools4RR/assets/lectures/06_org_eda_withnotes.pdf


## Installing Git/Github

http://r-pkgs.had.co.nz/git.html#git-init

## Git

<img class=center src=./git.png height=450>

http://git-scm.com/downloads

## Github

<img class=center src=./github.png height=450>

https://github.com/

## Github tour

https://github.com/jtleek/datasharing

## Pushing and pulling

<img class=center src=./scheme.png height=450>

http://gitready.com/beginner/2009/01/21/pushing-and-pulling.html

## Git on the command line practice

https://try.github.io/levels/1/challenges/1

## Git in Rstudio

<img class=center src=./inr2.png height=450>


## Git in Rstudio

<img class=center src=./inr3.png height=450>


## Git in Rstudio

<img class=center src=./inr4.png height=450>

## Git/Github practice

http://bit.ly/1H0qcwB
