---
title: "Tidying data"
author: "Jeff Leek"
date: "September 9, 2015"
output: ioslides_presentation
---

## A funny quote

> "Herein lies the dirty secret about most data scientists' work -- it's more data munging than deep learning. The best minds of my generation are deleting commas from log files, and that makes me sad. A Ph.D. is a terrible thing to waste."

http://adage.com/article/digitalnext/dear-madison-avenue-set-data-scientists-free/298676/

## Raw versus processed data

__Raw data__
* The original source of the data
* Often hard to use for data analyses
* Data analysis _includes_ processing
* Raw data may only need to be processed once

[http://en.wikipedia.org/wiki/Raw_data](http://en.wikipedia.org/wiki/Raw_data)

__Processed data__
* Data that is ready for analysis
* Processing can include merging, subsetting, transforming, etc.
* There may be standards for processing
* All steps should be recorded 

[http://en.wikipedia.org/wiki/Computer_data_processing](http://en.wikipedia.org/wiki/Computer_data_processing)


## 1,000 genomes example

```{r, cache=TRUE}
fileUrl = "ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_sample_info.xlsx"
library(downloader)
library(readr)
download(fileUrl,destfile="1000genomes.xlsx")

library(readxl)
# Have to skip one row because 
# there is an extra header
kg = read_excel("1000genomes.xlsx",
                    sheet=4,skip=1)
```


## Variable names

```{r}
head(kg,n=1)
head(kg$`Total Sequence`)
```



## strsplit

```{r}
strsplit(names(kg)," ")
strsplit(names(kg)[6]," ")
```

## gsub

```{r}
gsub(" ", "", names(kg))
```

## substr

```{r}
x = gsub(" ", "", names(kg))
substr(x,1,10)
```

## grep

```{r}
grep("Center",names(kg))
grepl("Center",names(kg))
```

## tolower and toupper

```{r}
tolower(names(kg))[1:5]
toupper(names(kg))[1:5]
```


## dplyr

<img class=center src=./dplyr.png height=350>


## Basically this whole part of the lecture from

<img class=center src=./stat545.png height=350>


https://stat545-ubc.github.io/block009_dplyr-intro.html



## gapminder example

```{r test1,warning=FALSE}
suppressPackageStartupMessages({library(dplyr)})
gd_url <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt"
gdf <- read.table(file = gd_url,sep="\t",header=T)
head(gdf)
str(gdf)
```

## tbl_df

```{r}
gtbl <- tbl_df(gdf)
gtbl
```

## glimpse

```{r}
glimpse(gtbl)
```

## filter

```{r}
filter(gtbl, lifeExp < 29)
filter(gtbl, country == "Rwanda")
```

## select

```{r}
select(gtbl,country,pop,continent)
```

## arrange

```{r}
arrange(gtbl,pop)
```


## desc

```{r}
arrange(gtbl,desc(pop))
```

## mutate

```{r}
gtbl = mutate(gtbl,newVar = (lifeExp / gdpPercap))
select(gtbl,lifeExp,gdpPercap,newVar)
```

## distinct

```{r}
distinct(gtbl)
```

## Key principle

<img class=center src=./bigsmall1.png height=350>

https://twitter.com/elliemcdonagh/status/469184554549248000

## Key principle

<img class=center src=./bigsmall2.png height=350>

https://twitter.com/elliemcdonagh/status/469184554549248000

## sample_n / sample_frac

```{r}
sample_n(gtbl,3)
```

## pipes

<img class=center src=./pipes.png height=350>


http://cran.r-project.org/web/packages/magrittr/index.html

## What does pipe do?

```{r}
gtbl %>% head
```

## What does pipe do? 

```{r}
gtbl %>% head(3)
```

## Example

Show me a random sample of the data for Asian countries with life expectancy < 65.

```{r}
gtbl1 = gtbl[gtbl$continent=="Asia",]
gtbl2 = gtbl1[gtbl1$lifeExp < 65,]
gtbl3 = gtbl2[sample(1:dim(gtbl2)[1],size=10),]
gtbl3
```

## w/pipes

```{r}
gtbl %>% filter(continent == "Asia") %>% 
         filter(lifeExp < 65) %>%
         sample_n(10)
```

## Another example

What is the average life expectancy by continent?

```{r}
gtbl %>% group_by(continent) %>% 
         summarize(aveLife = mean(lifeExp))
```

## The old-fashioned way

```{r}
tapply(gtbl$lifeExp,gtbl$continent,mean)
sapply(gtbl$continent,function(x){substr(x,1,3)})[1:10]
splitlife= split(gtbl$lifeExp,gtbl$continent)  
lapply(splitlife,mean)
```


## Common summarize options


<img class=center src=./summarize.png height=350>

## Passing to another argument

```{r}
x = rep(c(1,5),each=5)
z = x %>% rnorm(10,mean=.)
plot(z,pch=19,col="dodgerblue")
```

## Gotta be in the "Hadley mindset"

```{r, eval=FALSE}
sesmat = dst %>% group_by(coursera_user_id) %>% arrange(coursera_user_id,session_date) %>% 
  summarize(first_session = min(session_num),
            ses1 = c(1) %in% (session_num - first_session + 1),
            ses2 = c(2) %in% (session_num - first_session + 1),
            ses3 = c(3) %in% (session_num - first_session + 1),
            ses4 = c(4) %in% (session_num - first_session + 1),
            ses5 = c(5) %in% (session_num - first_session + 1),
            ses6 = c(6) %in% (session_num - first_session + 1),
            ses7 = c(7) %in% (session_num - first_session + 1),
            ses8 = c(8) %in% (session_num - first_session + 1),
            ses9 = c(9) %in% (session_num - first_session + 1),
            ses10 = c(10) %in% (session_num - first_session + 1),
            ses11 = c(11) %in% (session_num - first_session + 1),
            ses12 = c(12) %in% (session_num - first_session + 1))
```


## Merging data sets

The worst and most common task. 

```{r}
library(readxl)
kg_s4 = read_excel("1000genomes.xlsx",
                    sheet=4,skip=1)

kg_s1 = read_excel("1000genomes.xlsx",
                    sheet=1)
head(kg_s4,n=1)
head(kg_s1,n=1)
```

## names

```{r}
names(kg_s1)
names(kg_s4)
```

## merge

```{r}
kg_new = merge(kg_s4,kg_s1)
dim(kg_new)
names(kg_new)
```

## More on types of joins

<img class=center src=./sql-joins.jpg height=400>

https://stat545-ubc.github.io/bit001_dplyr-cheatsheet.html

## Superhero example

```{r}
superheroes <-
  c("    name, alignment, gender,         publisher",
    " Magneto,       bad,   male,            Marvel",
    "   Storm,      good, female,            Marvel",
    "Mystique,       bad, female,            Marvel",
    "  Batman,      good,   male,                DC",
    "   Joker,       bad,   male,                DC",
    "Catwoman,       bad, female,                DC",
    " Hellboy,      good,   male, Dark Horse Comics")

superheroes <- read.csv(text = superheroes, strip.white = TRUE)
```

https://stat545-ubc.github.io/bit001_dplyr-cheatsheet.html

## Superheroes example

```{r}
publishers <- 
  c("publisher, yr_founded",
    "       DC,       1934",
    "   Marvel,       1939",
    "    Image,       1992")
publishers <- read.csv(text = publishers, strip.white = TRUE)
```

## Inner join

```{r}
ijsp = inner_join(superheroes, publishers)
ijsp
```

## Comparison

<img class=center src=./innerjoin.png height=200>

## Left join

```{r}
ljsp = left_join(superheroes, publishers)
ljsp
```

## Comparison

<img class=center src=./leftjoin.png height=200>

## Outer join (using merge)

```{r}
ojsp = merge(superheroes, publishers, all = TRUE)
ojsp
```

## Merging different names

```{r}
superheroes = mutate(superheroes,
                jefflikes = (publisher=="Marvel"))
publishers = mutate(publishers,
                    jeff = (publisher == "Marvel"))
ij2 = inner_join(superheroes,publishers)
ij2
```


## Merging different names

```{r}
ij2 = inner_join(superheroes,publishers,
                    by=c("publisher"="publisher",
                            "jefflikes"="jeff"))

ij2
```

## Matching variables

```{r}
match(superheroes$publisher,publishers$publisher)
match(publishers$publisher,superheroes$publisher)
matchIndex = match(superheroes$publisher,publishers$publisher)
publishers[matchIndex,]
```


## Regular expressions

- Regular expressions can be thought of as a combination of literals and _metacharacters_
- To draw an analogy with natural language, think of literal text forming the words of this language, and the metacharacters defining its grammar
- Regular expressions have a rich set of metacharacters


## Regular expressions

Find all "BGI"
```{r}
library(readxl)
kg = read_excel("1000genomes.xlsx",sheet=4,skip=1)
table(kg$Center)
```

## grep

```{r}
grep("BCM",kg$Center)[1:10]
grepl("BCM",kg$Center)[1:10]
```

## stringr

```{r}
library(stringr)
str_detect(kg$Center,"BCM")[1:40]
str_subset(kg$Center,"BCM")[64:73]
```

https://github.com/hadley/stringr/blob/master/vignettes/stringr.Rmd
https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html

## A few string operations I use a lot

```{r}
str_trim("Jeff      ")
paste0("Jeffrey","Leek")
nchar("Jeffrey Leek")
```


## Literals: nuclear

<img class=center src=./nuclear.png height=400>


## Literals: Obama

<img class=center src=./obama.png height=400>


## But text is more complicated

We need a way to express
- whitespace word boundaries
- sets of literals
- the beginning and end of a line
- alternatives (“war” or “peace”) 

## Beginning of line with \^

```{r}
x = c("i think we all rule for participating",
"i think i have been outed",
"i think this will be quite fun actually",
"it will be fun, i think")

str_detect(x, "^i think")
```

## End of line with \$

```{r}
x = c("well they had something this morning",
"then had to catch a tram home in the morning",
"dog obedience school in the morning",
"this morning I'll go for a run")

str_detect(x, "morning$")
```

## Character lists with []

```{r}
x = c('Name the worst thing about Bush!',
'I saw a green bush',
'BBQ and bushwalking at Molonglo Gorge',
'BUSH!!')

str_detect(x,"[Bb][Uu][Ss][Hh]")
```

## Sets of letters or numbers

```{r}
x = c('7th inning stretch',
'2nd half soon to begin. OSU did just win.',
'3am - cant sleep - too hot still.. :(',
'5ft 7 sent from heaven')

str_detect(x,"^[0-9][a-zA-Z]")
```


## Negative classes

```{r}
x = c('are you there?',
'2nd half soon to begin. OSU did just win.',
'6 and 9',
'dont worry... we all die anyway!')

str_detect(x,"[^?.]$")
```

## . means anything

```{r}
x = c('its stupid the post 9-11 rules',
'NetBios: scanning ip 203.169.114.66',
'Front Door 9:11:46 AM',
'Sings: 0118999881999119725...3 !')

str_detect(x,"9.11")
```

## | means or

```{r}
x = c('Not a whole lot of hurricanes.',
'We do have floods nearly every day', 
'hurricanes swirl in the other direction',
'coldfire is STRAIGHT!')

str_detect(x,"flood|earthquake|hurricane|coldfire")
```

## More Metacharacters: ( and )

Subexpressions are often contained in parentheses to constrain the alternatives

```markdown
^([Gg]ood|[Bb]ad)
```

will match the lines

```markdown
bad habbit
bad coordination today
good, becuase there is nothing worse than a man in kinky underwear
Badcop, its because people want to use drugs
Good Monday Holiday
Good riddance to Limey
```

## More Metacharacters: ?

The question mark indicates that the indicated expression is optional

```markdown
[Gg]eorge( [Ww]\.)? [Bb]ush
```

will match the lines

```markdown
i bet i can spell better than you and george bush combined
BBC reported that President George W. Bush claimed God told him to invade I
a bird in the hand is worth two george bushes
```


## One thing to note...

In the following

```markdown
[Gg]eorge( [Ww]\.)? [Bb]ush
```

we wanted to match a “.” as a literal period; to do that, we had to “escape” the metacharacter, preceding it with a backslash In general, we have to do this for any metacharacter we want to include in our match


## More metacharacters: * and +

The * and + signs are metacharacters used to indicate repetition; * means “any number, including none, of the item” and + means “at least one of the item”

```markdown
(.*)
```

will match the lines

```markdown
anyone wanna chat? (24, m, germany)
hello, 20.m here... ( east area + drives + webcam )
(he means older men)
()
```


## More metacharacters: * and +

The * and + signs are metacharacters used to indicate repetition; * means “any number, including none, of the item” and + means “at least one of the item”

```markdown
[0-9]+ (.*)[0-9]+
```

will match the lines

```markdown
working as MP here 720 MP battallion, 42nd birgade
so say 2 or 3 years at colleage and 4 at uni makes us 23 when and if we fin
it went down on several occasions for like, 3 or 4 *days*
Mmmm its time 4 me 2 go 2 bed
```

## More metacharacters: { and }

{ and } are referred to as interval quantifiers; the let us specify the minimum and maximum number of matches of an expression

```markdown
[Bb]ush( +[^ ]+ +){1,5} debate
```

will match the lines

```markdown
Bush has historically won all major debates he’s done.
in my view, Bush doesn’t need these debates..
bush doesn’t need the debates? maybe you are right
That’s what Bush supporters are doing about the debate.
Felix, I don’t disagree that Bush was poorly prepared for the debate.
indeed, but still, Bush should have taken the debate more seriously.
Keep repeating that Bush smirked and scowled during the debate
```


## More metacharacters: and

- m,n means at least m but not more than n matches 
- m means exactly m matches
- m, means at least m matches

## More metacharacters: ( and ) revisited

- In most implementations of regular expressions, the parentheses not only limit the scope of alternatives divided by a “|”, but also can be used to “remember” text matched by the subexpression enclosed
- We refer to the matched text with \1, \2, etc.


## More metacharacters: ( and ) revisited

So the expression

```markdown
+([a-zA-Z]+) +\1 +
```

will match the lines

```markdown
time for bed, night night twitter!
blah blah blah blah
my tattoo is so so itchy today
i was standing all all alone against the world outside...
hi anybody anybody at home
estudiando css css css css.... que desastritooooo
```


## detecting phone numbers

```{r}
x = c('206-555-1122','206-332','4545','test')

phone = "([2-9][0-9]{2})[- .]([0-9]{3})[- .]([0-9]{4})"

str_detect(x,phone)

```


## Can this get ridiculous?

_You bet!_

<img class=center src=./regexgolf.png height=350>

https://regex.alf.nu/


## Like really ridiculous


<img class=center src=./regexgolf.png height=350>

https://xkcd.com/1313/



## A nice tutorial

<img class=center src=./stat545-regex.png height=350>

http://stat545-ubc.github.io/block022_regular-expression.html

## Important points about text in data sets

* Names of variables should be 
  * All lower case when possible
  * Descriptive (Diagnosis versus Dx)
  * Not duplicated
  * Not have underscores or dots or white spaces
* Variables with character values
  * Should usually be made into factor variables (depends on application)
  * Should be descriptive (use TRUE/FALSE instead of 0/1 and Male/Female versus 0/1 or M/F)

## Starting simple

```{r}
d1 = date()
d1
class(d1)
```

## Date class

```{r sysDate}
d2 = Sys.Date()
d2
class(d2)
```


## Formatting dates

`%d` = day as number (0-31), `%a` = abbreviated weekday,`%A` = unabbreviated weekday, `%m` = month (00-12), `%b` = abbreviated month,
`%B` = unabbrevidated month, `%y` = 2 digit year, `%Y` = four digit year

```{r ,dependson="sysDate"}
format(d2,"%a %b %d")
```

## Creating dates

```{r}
x = c("1jan1960", "2jan1960", "31mar1960", "30jul1960"); z = as.Date(x, "%d%b%Y")
z
z[1] - z[2]
as.numeric(z[1]-z[2])
```


## Converting to Julian 

```{r,dependson="sysDate"}
weekdays(d2)
months(d2)
julian(d2)
```



## Lubridate 

```{r lub}
library(lubridate); ymd("20140108")
mdy("08/04/2013")
dmy("03-04-2013")
```

[http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/](http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/)


## Dealing with times

```{r ,dependson="lub"}
ymd_hms("2011-08-03 10:15:03")
ymd_hms("2011-08-03 10:15:03",tz="Pacific/Auckland")
?Sys.timezone
```

[http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/](http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/)


## Some functions have slightly different syntax

```{r, dependson="lub"}
x = dmy(c("1jan2013", "2jan2013", "31mar2013", "30jul2013"))
wday(x[1])
wday(x[1],label=TRUE)
```


## Notes and further resources

* More information in this nice lubridate tutorial [http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/](http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/)
* The lubridate vignette is the same content [http://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html](http://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html)
* Ultimately you want your dates and times as class "Date" or the classes "POSIXct", "POSIXlt". For more information type `?POSIXlt`


## Lab time

* [dplyr](https://github.com/SISBID/Module1/blob/gh-pages/labs/dplyr-lab.Rmd)
* [regex](https://github.com/SISBID/Module1/blob/gh-pages/labs/regex-lab.Rmd)
* [merging](https://github.com/SISBID/Module1/blob/gh-pages/labs/merging-lab.Rmd)
* [final](https://github.com/SISBID/Module1/blob/gh-pages/labs/final-lab.Rmd)


