---
title: "Tidying data"
author: "Jeff Leek"
date: "September 9, 2015"
output: ioslides_presentation
---

## A funny quote

> "Herein lies the dirty secret about most data scientists' work -- it's more data munging than deep learning. The best minds of my generation are deleting commas from log files, and that makes me sad. A Ph.D. is a terrible thing to waste."

http://adage.com/article/digitalnext/dear-madison-avenue-set-data-scientists-free/298676/

## Raw versus processed data

__Raw data__
* The original source of the data
* Often hard to use for data analyses
* Data analysis _includes_ processing
* Raw data may only need to be processed once

[http://en.wikipedia.org/wiki/Raw_data](http://en.wikipedia.org/wiki/Raw_data)

__Processed data__
* Data that is ready for analysis
* Processing can include merging, subsetting, transforming, etc.
* There may be standards for processing
* All steps should be recorded 

[http://en.wikipedia.org/wiki/Computer_data_processing](http://en.wikipedia.org/wiki/Computer_data_processing)


## 1,000 genomes example

```{r, cache=TRUE}
fileUrl = "ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_sample_info.xlsx"
library(downloader)
library(readr)
download(fileUrl,destfile="1000genomes.xlsx")

library(readxl)
# Have to skip one row because 
# there is an extra header
kg = read_excel("1000genomes.xlsx",
                    sheet=4,skip=1)
```


## Variable names

```{r}
head(kg,n=1)
head(kg$`Total Sequence`)
```



## strsplit

```{r}
strsplit(names(kg)," ")
strsplit(names(kg)[6]," ")
```

## gsub

```{r}
gsub(" ", "", names(kg))
```

## substr

```{r}
x = gsub(" ", "", names(kg))
substr(x,1,10)
```

## grep

```{r}
grep("Center",names(kg))
grepl("Center",names(kg))
```

## tolower and toupper

```{r}
tolower(names(kg))[1:5]
toupper(names(kg))[1:5]
```


## dplyr

<img class=center src=./dplyr.png height=350>


## Basically this whole part of the lecture from

<img class=center src=./stat545.png height=350>


https://stat545-ubc.github.io/block009_dplyr-intro.html



## gapminder example

```{r test1,warning=FALSE}
suppressPackageStartupMessages({library(dplyr)})
gd_url <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt"
gdf <- read.table(file = gd_url,sep="\t",header=T)
head(gdf)
str(gdf)
```

## tbl_df

```{r}
gtbl <- tbl_df(gdf)
gtbl
```

## glimpse

```{r}
glimpse(gtbl)
```

## filter

```{r}
filter(gtbl, lifeExp < 29)
filter(gtbl, country == "Rwanda")
```

## select

```{r}
select(gtbl,country,pop,continent)
```

## arrange

```{r}
arrange(gtbl,pop)
```


## desc

```{r}
arrange(gtbl,desc(pop))
```

## mutate

```{r}
gtbl = mutate(gtbl,newVar = (lifeExp / gdpPercap))
select(gtbl,lifeExp,gdpPercap,newVar)
```

## distinct

```{r}
distinct(gtbl)
```

## Key principle

<img class=center src=./bigsmall1.png height=350>

https://twitter.com/elliemcdonagh/status/469184554549248000

## Key principle

<img class=center src=./bigsmall2.png height=350>

https://twitter.com/elliemcdonagh/status/469184554549248000

## sample_n / sample_frac

```{r}
sample_n(gtbl,3)
```

## pipes

<img class=center src=./pipes.png height=350>


http://cran.r-project.org/web/packages/magrittr/index.html

## What does pipe do?

```{r}
gtbl %>% head
```

## What does pipe do? 

```{r}
gtbl %>% head(3)
```

## Example

Show me a random sample of the data for Asian countries with life expectancy < 65.

```{r}
gtbl1 = gtbl[gtbl$continent=="Asia",]
gtbl2 = gtbl1[gtbl1$lifeExp < 65,]
gtbl3 = gtbl2[sample(1:dim(gtbl2)[1],size=10),]
gtbl3
```

## w/pipes

```{r}
gtbl %>% filter(continent == "Asia") %>% 
         filter(lifeExp < 65) %>%
         sample_n(10)
```

## Another example

What is the average life expectancy by continent?

```{r}
gtbl %>% group_by(continent) %>% 
         summarize(aveLife = mean(lifeExp))
```

## Common summarize options



<img class=center src=./summarize.png height=350>

